---
title: "Chapter 10: Correlation"
author: "Paul Northrop"
date: ""
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 10: Correlation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat0002.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE,
                      fig.width = 7, fig.height = 5, fig.align = 'center')
knitr::opts_knit$set(global.par = TRUE)

required <- c("anscombiser")
if (!all(unlist(lapply(required, function(pkg) requireNamespace(pkg, quietly = TRUE)))))
  knitr::opts_chunk$set(eval = FALSE)
```

This vignette provides some R code that is related to some of the content of [Chapter 10 of the STAT0002 notes](https://paulnorthrop.github.io/stat0002book/correlationchapter.html#correlationchapter) correlation.  

```{r, message = FALSE}
library(stat0002)
```

```{r, echo = FALSE}
par(mar = c(4, 4, 1, 1))
```

## Example data

We use the exchange rate data described at the start of [Section 10.1](https://paulnorthrop.github.io/stat0002book/correlationchapter.html#correlation-a-measure-of-linear-association) of the notes. These data are available in the data frame `exchange`. We look at the first 6 rows of the data and the last 6 rows and produce plots like Figures 10.1 and 10.2 of the notes.

```{r}
head(exchange)
tail(exchange)
# Figure 10.1
plot(exchange, pch = 16, xlab = "Pounds sterling vs. US dollars",
     ylab = "Pounds sterling vs. Canadian dollars", bty = "l", 
     main = "Raw data")
# Calculate the log-returns
USDlogr <- diff(log(exchange$USD.GBP))
CADlogr <- diff(log(exchange$CAD.GBP))
# Figure 10.2
plot(USDlogr, CADlogr, pch = 16, xlab = "Pounds sterling vs. US dollars",
     ylab = "Pounds sterling vs. Canadian dollars", bty = "l", 
     main = "Log-returns" )
```


## The R function `cor`

R provides a function `cor` to calculate sample correlation coefficients.  By default, it calculates the sample (product moment) correlation coefficient given in [Section 10.2.1](https://paulnorthrop.github.io/stat0002book/correlationchapter.html#estimation). 

We can supply either two vectors `x` and `y` of equal length or a matrix with columns containing the vectors of data for which we want to calculate the sample correlation coefficient. In this case, we create a 2-column matrix using `cbind` function to combine the 2 vectors `USDlogr` and `CADlogr` columnwise into a matrix. 

If we supply a matrix then R will return a matrix containing the sample correlation coefficients between all possible pairs of columns in the input matrix, including between each column and itself, which produces values of 1 on the diagonal of the output matrix. If we supply vectors `x` and `y` then it does not matter which vector is entered as `x` and which as `y`.   

```{r}
cor(x = USDlogr, y = CADlogr)
cor(x = CADlogr, y = USDlogr)
cor(x = cbind(USDlogr, CADlogr))
```

## Anscombe's Quartet of datasets

In [Section 10.3.6](https://paulnorthrop.github.io/stat0002book/correlationchapter.html#anscombes-datasets) Anscombe's Quartet of dataset are provided in Figure 10.9. These 4 datasets are available as separate data frames in the `anscombiser` package @anscombiser.  Use `install.packages("anscombiser")` to install this package.

```{r}
library(anscombiser)
```

```{r}
cor(anscombe1)
cor(anscombe2)
cor(anscombe3)
cor(anscombe4)
```

We see that the paired data in these datasets have almost exactly the same sample correlation coefficient.  They have many other summary statistics in common.  See [Section 10.3.6](https://paulnorthrop.github.io/stat0002book/correlationchapter.html#anscombes-datasets) for details.

The `method` argument to `cor` enables us to ask R to calculate Spearman's rank correlation coefficient (See [Section 2.3.5](https://paulnorthrop.github.io/stat0002book/descriptive.html#corr1)). The values of this sample correlation coefficient differ between these datasets.

```{r}
cor(anscombe1, method = "spearman")
cor(anscombe2, method = "spearman")
cor(anscombe3, method = "spearman")
cor(anscombe4, method = "spearman")
```


## References

<script type="text/x-mathjax-config">
   MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 125, availableFonts: [] }  });
</script>
