---
title: "Creating PDF files using R markdown"
author: "Paul Northrop"
date: ""
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Creating PDF files using R markdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat0002.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE)
```

STAT0004: R Markdown tutorial at the end of Term 1.

[R Markdown from RStudio](https://rmarkdown.rstudio.com/).

Helper wrapper files in `stat0002`.

* `stat0002::word()`.
* `stat0002::html()`

Command line using `accessr` or using the Knit button.

Look at the STAT0004/STAT0006 Moodle pages

# Cut-and-paste using a Word Processor

This is the least elegant and most time-consuming approach, but it can be useful ...

# PDF

Accessibility (links). May not matter for your personal use.

# Via Word using `accessr`

## Windows users

```{r}
library(accessr)
```

Either 

* `OfficeToPDF` works only on a Windows operating system.

or following the general instructons for Mac / linux users below.

## Mac / linux users

* Create the Word file using `rmd2word()`.
* (Perhaps convert it to an alternative format.)
* Open it in a Word Processor and print to PDF. 


# Via HTML using Google print

* PDF file is not accessible

# Directly, using LaTeX

* Requires a LaTeX installation
* PDF file is not accessible



## References

<script type="text/x-mathjax-config">
   MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 125, availableFonts: [] }  });
</script>
