---
title: "Chapter 6: QQ plots"
author: "Paul Northrop"
date: ""
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 6: QQ plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat0002.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE,
                      fig.width = 6, fig.height = 6, fig.align = 'center')
knitr::opts_knit$set(global.par = TRUE)
```

```{r, include = FALSE}
# Set margins globally
par(mar = c(4, 4, 1, 1))
```

This vignette produces some of the QQ plots that appear at the end of [Chapter 6](https://paulnorthrop.github.io/stat0002book/simple.html#simple) of the notes.

```{r, warning = FALSE, message = FALSE}
library(stat0002)
```

## Normal QQ plots

We create plots that are similar to those in Figure 6.24 of the notes. They are slightly different from the ones in the notes because R's `qqnorm` function uses a method of calculating the theoretical quantiles that is designed specifically for the normal distribution, whereas in [Section 6.11.1](https://paulnorthrop.github.io/stat0002book/simple.html#normal-qq-plots) we used an approach that ties in with the way that we had calculated sample quantiles earlier in STAT0002.  The dashed lines in the plots are drawn through the (sample and theoretical) lower and upper quartiles.

```{r, echo = FALSE}
set.seed(39)
```

```{r, fig.show='hold', out.width='47%', fig.align='default'}
# Normal
normal <- rnorm(100)  
qqnorm(normal, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "normal")
qqline(normal, lwd = 3, lty = 2)
# Create an outlier
outlier <- normal
outlier[100] <- 6
qqnorm(outlier, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "outlier")
qqline(outlier, lwd = 3, lty = 2)
```

```{r, echo = FALSE}
set.seed(39)
```

```{r, fig.show='hold', out.width='47%', fig.align='default'}
# Heavy-tailed (Student's t distribution, with 2 degrees of freedom)
StudentsT <- rt(100, 2)
qqnorm(StudentsT, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "heavy-tailed")
qqline(StudentsT, lwd = 3, lty = 2)
# Light-tailed (Uniform(-3/4, 3/4))
uniform <- runif(100, -0.75, 0.75)
qqnorm(uniform, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "light-tailed")
qqline(uniform, lwd = 3, lty = 2)
```

```{r, fig.show='hold', out.width='47%', fig.align='default'}
# Positively skewed (gamma(2, 1) distribution)
gam <- rgamma(100, shape = 2)
qqnorm(gam, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "positive skew")
qqline(gam, lwd = 3, lty = 2)
# Negatively skewed (10 - gamma(2, 1) distribution)
neggam <- 10 - rgamma(100, 2)
qqnorm(neggam, pch = 16, xlab = "Theoretical N(0, 1) Quantiles", main = "negative skew")
qqline(neggam, lwd = 3, lty = 2)
```

## Exponential QQ plots

We produce an exponential QQ plot based on the waiting times between births in the Australian birth times dataset. The rate $\lambda$ of the assumed Poisson process of births is estimated using the reciprocal of the sample mean, as detailed in the notes.  The plot is equivalent to the figure in the notes, but the simulation envelopes are different because, unless we fix the random number seed, the simulated datasets on which the envelopes are based will be different each time we call `qexp`.

```{r}
# Calculate the waiting times until each birth
waits <- diff(c(0, aussie_births[, "time"]))
# Produce the QQ plot
lambdahat <- qqexp(waits, envelopes = 19)
# Estimate of lambda: rate of births per hour
lambdahat * 60
```

The `qqexp` function offers the option to estimate $\lambda$ using $\ln 2 / m$, where $m$ is the sample median of the waiting times. **Can you see why this makes sense?**  The following code implements this and shows how we can alter the appearance of the plot.

```{r}
lambdahat2 <- qqexp(waits, statistic = "median", envelopes = 19, pch = 16,
                    line = list(lty = 2, lwd = 2, col= "blue"))
# Estimate of lambda calculated from the sample median
lambdahat2 * 60
```

<script type="text/x-mathjax-config">
   MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 125, availableFonts: [] }  });
</script>
