---
title: "Chapter 3: Probability"
author: "Paul Northrop"
date: ""
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 3: Probability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat0002.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE,
                      fig.width = 7, fig.height = 5, fig.align = 'center')
```

This vignette provides R code to reproduce some of the content of [Chapter 3 of the STAT0002 notes](https://paulnorthrop.github.io/stat0002book/probability.html).

## Kerrich's coin data

A summary of these data are available in the data frame `kerrich`.  Use `?kerrich` for more information.

```{r, message = FALSE}
library(stat0002)
```

```{r, kerrichplot}
# This code produces the plot in Figure 3.1 of the STAT0002 notes
plot(kerrich$throws, kerrich$heads / kerrich$throws,
     ylab = "proportion of heads",
     xlab = "number of throws (logarithmic scale)", lwd = 2, type = "l",
     log = "x", ylim = c(0,1), axes = FALSE)
abline(h = 0.5, lty = 2)
axis(1, labels = as.character(c(3, 10, 30, 100, 300, 1000, 3000, 10000)),
     at=c(3, 10, 30, 100, 300, 1000, 3000, 10000))
axis(2, labels = c(0, 0.2, 0.4, 0.5, 0.6, 0.8, 1.0),
     at=c(0, 0.2, 0.4, 0.5, 0.6, 0.8, 1.0))
```

We could extract from the data frame `kerrich` the number of trials (the 10,000 throws of the coin) and the number of heads.  From this we can calculate an estimate of the probability $P(H)$ that the outcome of a trial is a head.

```{r, estimatep}
trials <- kerrich[nrow(kerrich), "throws"]
heads <- kerrich[nrow(kerrich), "heads"]
c(heads, trials)
phat <- heads / trials
phat
```

## Graduate Admissions at Berkeley

The object `berkeley` is a 3-dimensional array that contains information about applicants to graduate school at UC Berkeley in 1973 for the six largest departments. Use `?UCBAdmissions` for more information. The 3 dimensions of the array correspond to the gender of the applicant (dimension named `Gender`), whether or not they were admitted (named `Admit`) and a letter code for the department to which they applied (named `Dept`).  A given entry in `berkeley` gives the total number of applicants in the corresponding (`Admit`, `Gender`, `Dept`) category. In Chapter 3 we consider only the dimensions `Admit` and `Gender`.  

The following code collapses the 3-dimensional array to a 2-dimensional array by summing the frequencies over all the six departments, which gives the central part of Table 3.3 in the notes. 

```{r berk2way}
# 2-way table: sex and outcome
sex_outcome <- apply(berkeley, 2:1, FUN = sum)
colnames(sex_outcome) <- c("A", "R")
rownames(sex_outcome) <- c("M", "F")
sex_outcome
```

Now we add the column and row totals.

```{r berk2wayplustotals}
# Add column totals
sex_outcome <- rbind(sex_outcome, total = colSums(sex_outcome))
# Add row totals
sex_outcome <- cbind(sex_outcome, total = rowSums(sex_outcome))
sex_outcome
```

We can divide by $n$ (4526) to produce Table 3.5 of the notes.

```{r berkprobs}
# Convert frequencies to probabilities
pso <- sex_outcome/ sex_outcome[3, 3]
round(pso, 3)
```

Notice that `sex_outcome[3, 3]` extracts the [3, 3] element of the matrix `sex_outcome`, that is, 4526.

**Can you use R to perform some of the calculations that are performed in Section 3.4 and/or 3.5 of the notes?**

## Blood types data

A summary of these data are available in the data frame `blood_types`.

```{r bloodtypes}
blood_types
```

One way to estimate the probabilities given on which Table 3.13 is based is to use the `aggregate` function.

```{r bloodcalcs}
aggregate(percentage ~ Rh, data = blood_types, FUN = sum) 
aggregate(percentage ~ ABO, data = blood_types, FUN = sum)
```

To divide by 100 to convert the frequencies to estimates of probability we could do the following.

```{r bloodcalcs2}
propn <- function(x) sum(x) / 100
aggregate(percentage ~ Rh, data = blood_types, FUN = propn) 
aggregate(percentage ~ ABO, data = blood_types, FUN = propn)
```


<script type="text/x-mathjax-config">
   MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 125, availableFonts: [] }  });
</script>
