---
title: "Template Rmd document for Exercises 1"
author: "Paul Northrop"
date: "`r format(Sys.time(), '%d %B, %Y')`"
lang: "en-gb"
output: 
  html_document: default
  word_document: default
---

```{r, include = FALSE}
# Set some default options for the R Markdown code (called "chunks") below
# Each R code chunk starts with ```{r name_of_chunk} and ends with ```
#   echo = TRUE: the code in the chunk will be printed
#   prompt = TRUE: the R prompt > will appear with R code
#   fig.align = 'center': plots will be centered on the page
#   message = FALSE and warning = FALSE turns off messages and warnings

# The chunk below that starts ```{asis, echo = FALSE} and ends ``` provides a
# way to include text comments in an Rmd document that will not appear in the 
# output.
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, fig.align = 'center',
                      message = FALSE, warning = FALSE)
```

```{asis, echo = FALSE}
Of course, you should replace the text and R code below with your own work.

The R Markdown chunk immediately below loads the `stat0002` package. If you do not have this installed then change `eval = TRUE` to `eval = FALSE` (`eval` is short for evaluate).
```

```{r load_stat0002, eval = TRUE, echo = FALSE}
library(stat0002)
```

```{r, echo = TRUE, eval = TRUE, echo = FALSE}
dat <- "https://github.com/paulnorthrop/stat0002/blob/master/days.Rdata?raw=true"
load(url(dat))
```

# 1 

Perform calculations by hand or use R. If you do the calculations by hand then given your final answer in the form: (minimum, lower quartile, median, upper quartile, maximum). If you use R then include your code and the output that it produces.

# 2

The following code produces a boxplot that is similar to the one in the Exercises 1 instructions.

```{r boxplot}
box_plot(days, horizontal = TRUE, 
         xlab = "mystery variable (days)",
         axes = TRUE, cex.lab = 1.5, cex.axis = 1.5)
```
# 3

Use `?hist` to see instructions for creating a histogram.

```{r}
?hist
```

# 4

Use `?stem` to see instructions for creating a stem-and-leaf plot.

# 5

Write text here either to answer question 5.

# 6

Write text here either to answer question 6.
