---
title: "A basic example Rmd document"
author: "Paul Northrop"
date: "`r format(Sys.time(), '%d %B, %Y')`"
lang: "en-gb"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, fig.align = 'center')
knitr::opts_knit$set(global.par = TRUE)
```

We load the `stat0002` R package.

```{r stat0002, warning = FALSE, message = FALSE}
library(stat0002)
```

We simulate some example data and store it in the vector `x`.

```{r example_data}
x <- rnorm(100)
```

# Summary statistics

```{r}
five_number(x)
summary(x)
```

See the [Summary statistics section](https://paulnorthrop.github.io/stat0002book/descriptive.html#summary-statistics) of the STAT0002 notes. See also the `stat0002` R package article [Chapter 2: Descriptive Statistics](https://paulnorthrop.github.io/stat0002/articles/stat0002-ch2a-descriptive-statistics-vignette.html) to help you consider why some of these numbers seem to be in disagreement.

# Plots

```{r plot_margins, echo = FALSE}
# We adjust the plot margins.
oldpar <- par(mar = c(4, 4, 0, 0))
```

```{r Rfigure, fig.alt = "Alt text to describe the content of the figure.", fig.cap = "A boxplot of a random sample from a standard normal distribution.", echo = TRUE, fig.width = 6, fig.height= 4, fig.align='center'}
box_plot(x, horizontal = TRUE, xlab = "Sample values")
```

See the [Boxplots](https://paulnorthrop.github.io/stat0002book/descriptive.html#boxplots) of the STAT0002 notes. See also the `stat0002` R package article [Chapter 2: Graphs (One Variable)](https://paulnorthrop.github.io/stat0002/articles/stat0002-ch2b-graphs-vignette.html) to help you consider why some of these numbers seem to be in disagreement.

```{r plot_margins_reset, echo = FALSE}
# Reset the plot margins to their origin values.
par(oldpar)
```

# Lists

Bullet points

* Bullet 1
* Bullet 2

Numbered points

1. Point 1
2. Point 2
3. Point 3

#  Mathematics

We can typeset mathematics within sentences by place mathematical notation inside a pair of dollar signs \$ \$, for example, $ax^2 + b x + c$, $\log_e y$, $\ln y$, $\exp(x)$, $\frac{1}{n}$ or $1/n$.

## Displayed equations

$$y = mx + c.$$

$$
\begin{aligned}
P(X \leq x)
&= P(F^{-1}(U) \leq x) \\
&= P(F(F^{-1}(U)) \leq F(x)) \\
&= P(U \leq F(x)) = F(x).
\end{aligned}
$$

Tips (prompted by issues when creating Word output):

* We need to wrap LaTeX maths environments in ```$$ ... $$```.  Otherwise, the maths is ignored.
* Use ```\text{}``` or ```\mbox{}``` for text in an maths environment. (```\rm{}``` causes problems.)
